---
title: "Small-plot_analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Small-plot_analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(CBADASReml)
library(asreml)
```

# CBADASReml

## Introduction

A typical type of experimental analysis that this package will assist with is
the small plot trial. The trial design is not handled at all by this package,
and we recommend the use of \pkg{FielDHub} for creating small plot trial
designs.

After gathering data from the experiment, it is useful to perform an
exploratory data analysis on the raw data, and for this you will likely use a
combination of \pkg{tidyverse} and \pkg{ggplot2} for this, to visualise
outliers, the data distribution, etc. in order to get a preliminary grasp of
what you are working with, and to observe the data for any issues that will
hinder the actual analysis.

Analysis of small plot trials is done using linear mixed-effects models (LMMs),
using \pkg{asreml} or \pkg{glmmTMB}, and this package contains helper functions
for this type of model.

## Typical Workflow

The following will outline a typical workflow starting from an existing
univariate LMM.

### LMMs and ASReml

Let:

\begin{equation*}
y
=
X b
+
Z u
+
e
\end{equation*}

Where

- $y \in \mathbb{R}^{n}$ is the vector of observations.
- $b \in \mathbb{R}^{p}$ is the vector of fixed effects.
- $X \in \mathbb{R}^{n \times p}$ is the design matrix for the fixed effects.
- $u \in \mathbb{R}^{q}$ is the vector of random effects.
- $Z \in \mathbb{R}^{n \times q}$ is the design matrix for the random effects.
- $e \in \mathbb{R}^{n}$ is the vector of residuals.

It is assumed that the vectors $u$ and $e$ are independent, and their joint
distribution is a multivariate Gaussian described by:

\begin{equation*}
\begin{bmatrix}
u \\
e
\end{bmatrix}
\sim
N \bigg(
  \begin{bmatrix}
  0 \\
  0
  \end{bmatrix}
  ,
  \begin{bmatrix}
  G & 0 \\
  0 & R
  \end{bmatrix}
\bigg)
\end{equation*}

Where $G$ and $R$ are covariance matrices for $u$ and $e$ and are functions of
the *variance parameters* $\sigma_g$ and $\sigma_r$, and finally the dependent
variable $y$ is described by:

\begin{equation*}
y
\sim
N \big(
X b
,
Z G Z^\intercal
+
R
\big)
\end{equation*}

These $G$ and $R$ covariance matrices, or *covariance structures*, are very
important. These represent the covariance of the random effects and the
residuals respectively, and $\sigma_g$ and $\sigma_r$, the *covariance
structure parameters* are the things we try and estimate when we use ASReml.

In the case of independent and identically distributed (IID) residual
structures, the residual covariance parameter $\sigma_r$ will be some *variance
component* $\sigma_e^2$.



We may use the R package \pkg{asreml} as follows to implement an LMM:

```{r}
mod <- asreml(
    fixed = yield ~ Variety + Nitrogen + Variety:Nitrogen,
    random = ~ idv(Blocks) + idv(Blocks):idv(Wplots),
    residual = ~ idv(units),
    data = oats,
    trace = FALSE
)
```

In this case we are using the `oats` data:

- `yield` is our $y$ vector.
- `Variety`, `Nitrogen` and their interaction `Variety:Nitrogen` are in the $X$
  matrix.
- `Blocks` and `Blocks:Wplots` are in the $Z$ matrix. We specify the `idv`
  variance model function, which is the homogeneous scaled identity variance
  structure, i.e. $\sigma^2 I$. In other words, the term is IID:
  + $u_b \sim N(0, \sigma_b^2 I_b)$
  + $u_{bp} \sim N(0, \sigma_{bp}^2 I_{bp})$
  + $u = [ \; u_b \; u_{bp} \; ]^\intercal$
  + $Z = [ \; Z_b \; Z_{bp} \; ]$ (the design matrices for `Blocks` and
    `Blocks:Wplots`)
  + Therefore, $G = \bigoplus_{i = 1}^2 G_i = \begin{bmatrix} \sigma_b^2 I_b &
    0 \\ 0 & \sigma_{bp}^2 I_{bp} \end{bmatrix}$
- `idv(units)` is the residual variance structure, in this case we are saying
  the residual term has the homogeneous covariance matrix: $\sigma^2 I$. This
  is basically saying all observations have the same error distribution (IID):
  + $e \sim N(0, \sigma_e^2 I_n)$
  + Therefore, $R = \sigma_e^2 I_n$, the simplest $R$ structure.

Upon executing the above in R, the ASReml-R code will attempt to *estimate* the
variance components $\sigma_b^2$, $\sigma_{bp}^2$, and $\sigma_e^2$, using the
REML algorithm.

## CBADASReml

The \pkg{CBADASReml} package has a few functions to make working with LMMs in R
a bit easier, with consistent outputs.

### `anova_table`

The `anova_table` function takes multiple \pkg{asreml} objects, and returns a
dataframe containing all of the model's ANOVA tables:

```{r}
test_data <- oats
test_data["yield2"] <- oats["yield"] * runif(nrow(oats["yield"]))

mod1 <- asreml(
    fixed = yield ~ Variety + Nitrogen + Variety:Nitrogen,
    random = ~idv(Blocks) + idv(Blocks):idv(Wplots),
    residual = ~idv(units),
    data = test_data,
    trace = FALSE
)
mod2 <- asreml(
    fixed = yield2 ~ Variety + Nitrogen + Variety:Nitrogen,
    random = ~idv(Blocks) + idv(Blocks):idv(Wplots),
    residual = ~idv(units),
    data = test_data,
    trace = FALSE
)

wald(mod1)

anova_table(mod1, mod2)
```

The above example uses the `oats` data included with the \pkg{asreml} package,
and creates an altered `yield` variable for the purpose of demonstration.

The function returns a dataframe, one row per fixed effect, and $n+1$ columns
for $n$ models. Each value is the p-value taken from the wald test performed by
`asreml::wald.asreml`.
